Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-06-27 Delaneyetal2024.tex
online edit

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom, which varies with water velocity and channel width.
  In subaerial channels, water discharge variations are accommodated  by flow depth and width changes, along with water velocity.
  However, in subglacial channels,  water discharge changes mainly result in velocity variations, as the channel geometry evolves slowly (over days).
  Here, we present relationships between water discharge and sediment transport under several channel types.
  They demonstrate that sediment transport is highly non-linear with respect to water discharge with fixed subglacial channel size, and in fact can very less that subaerial flow if water discharge changes slowly enough to be in equilibrium with subglacial channel size.
  We then apply hydrographs from an Alpine glacier and the Greenland Ice Sheet to subglacial and subaerial hydraulics models, revealing greater variability in the inferred sediment transport capacity in subglacial than in subaerial channels, as subglacial channels evolve in size.
  This variability stems from the slowly evolving channel size that forces water velocity to accommodate variations in water discharge.
  Furthermore, the changing channel size means that sediment transport capacity peaks prior to the maximum water discharge.
  Results also show that high subglacial sediment transport rates and water velocities can occur across a wide range of water discharges, meaning that in a transport-limited system there could be a variable relationship between sediment and water discharge.
  These findings are discussed in the context of understanding sediment discharge from glaciers with different hydro-climatic forcings and assumptions of channel shape.
  We also examine the implications for interpreting records of sediment discharge in glacierized catchments
\end{abstract}

@@

Periods of millennia or longer, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \citep[c.f.][]{hallet1979,iverson2012,ugelvig2018}.
Pressurized subglacial water can transport this sediment from underneath glaciers \citep{walder1994,creyts2013,beaud2018,delaney2019}, if it is reachable by the water.

@@

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.6\linewidth]{Fig1.pdf}
  \caption{Sketch for the different responses of subglacial and subaerial channels to increased water discharge over short time-scales.
    Arrow length denotes water velocity magnitudes in the subglacial (subaerial) $v_g$ ($v_f$) channels.
    $S_g$ ($S_f$) represents the wetted area in subglacial (subaerial) channels.
    The subglacial channel width $w_g$ remains unchanged, while the subaerial channel width $w_f$ evolves with water discharge.
    Subglacial (subaerial) shear stress $\tau_g$ ($\tau_f$) is responsible for the mobilization of sediment.
  }
  \label{fig:cartoon}
\end{figure}

Sediment mobilization in subaerial and subglacial channels, therefore, responds differently to changing water discharge.
These differences are implicitly included in a range of models quantifying sediment transport in both subglacial and subaerial channels  \citep[e.g.][]{walder1994,alley1997,tucker1997,creyts2013,beaud2018,delaney2019,hewitt2019,wickert2019}.
To date, explicit parameterizations of subglacial sediment transport capacity with respect to water discharge has assumed fixed channel size and demonstrates a very non-linear response in sediment transport capacity to water discharge \citep{alley1997}.
The modeling frameworks examining subglacial sediment transport with evolving channel size minimally discuss the impact of different hydrological forcing on sediment transport capacity with the aim of understanding the impact of hydro-climatic conditions on subglacial sediment dynamics \citep{creyts2013,beaud2018,delaney2019,hewitt2019}.
Yet, sediment discharge capacity controls the mobilization and deposition of sediment, making it one of the most fundamental aspects of subglacial sediment evacuation, especially over short timescales.

Despite the physical differences between the subaerial and subglacial systems, water discharge and sediment export are often compared in contemporary glacial environments \citep[e.g.][]{willis1996,pearce2003,richards2003,swift2005,chu2009,tedstone2012,chu2012,overeem2017,delaney2018,swift2021,lu2022,andresen2024}.
These comparisons are natural given that the majority of sediment evacuated from glaciers over contemporary seasonal timescales is done so fluvially \citep[e.g.][]{collins1989, riihimaki2005,delaney2024}.
In subaerial systems, water discharge and sediment transport capacity for a grain size are closely linked \citep[e.g. ][]{schmidt2006,pitlick2021} \ian{check citiation}.
However, the potential for a highly variable response in sediment transport capacity in subglacial sediment transport may affect the interpretation of sediment transport records in glacial systems \citep[e.g.][]{ganti2016,mancini2023}, if properties associate with subaerial systems are assumed.
This relationship should be clarified as sediment dynamics change in glacierized regions along with changing hydrology, increased variability of water discharge \citep{lane2019} and extreme glacier melt  \citep[e.g.][]{cremona2023,overeem2015} that may drive high magnitude sediment transport events.

This manuscript as two objectives: 1) evaluate the variability of subglacial sediment transport capacity with evolving subglacial channels size and different hydrographs and 2) establish if subseasonal variations in water discharge can be covary with sediment transport capacity in subglacial systems.
First, to reenforce the highly variable behavior of subglacial sediment transport capacity,  we present algebric formulations of the sediment transport capacity response to water discharge in subaerial, pipe-flow and equilibrium R-channel conditions, extending the relationships presented in \citet{alley1997}.
We then use numerical models to examine  sediment transport capacity in subglacial channels that evolve in size and subaerial channels.
They are then applied to proglacial hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and  a land-terminating glacier in Greenland (Leverett Glacier).
We run models with an ensemble of parameters to evaluate the sediment transport capacity variability in subglacial channels compared to subaerial ones.
Lastly, model outputs demonstrate the particular processes that can drive variability in sediment discharge capacity in subglacial systems.
Together, these findings indicate differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
The manuscript then uses the identified differences between the subaerial and subglacial channel types to discuss the implications for interpreting the processes driving variability in sediment transport records and the impact of these differences on sediment export from glaciers.

@@

\section{Methods}
\label{sect:meth}
The two models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table~\ref{table:vpm}).
Both models use the measured discharge to calculate water velocity, shear stress, and width-integrated shear stress, upon which both suspended sediment and bedload transport depend \citep[Figure \ref{fig:cartoon}; ][]{shields1936}.
Our evaluation in terms of shear stress omits the selection of a sediment transport relationship and a grain-size parameter \citep[e.g.][]{shields1936,meyer1948}.
Through this choice we also negate sorting processes

@@

Here, it is assumed that the water is transported through a subglacial channel \citep[Figure~\ref{fig:cartoon}; ][]{rothlisberger1972} beneath a glacier with channel length $l$, with a flat bed and a mean thickness of $h_{ice}$.
The channel  size grows from melt due to frictional heating from water flow and closes due to the creep  of the ice.
The formulation here does not consider the englacial storage of water.
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_1 \frac{Q \Delta h}{l} - C_2 \left(h_{o}-\frac{\Delta h}{2}\right)^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent where $t$ is time, $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (complete values given in Table~\ref{table:vpm}), $g$ is the acceleration due to gravity, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change over $l$, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the mean ice overburden pressure expressed in meter water equivalent ($\rho_w$ is density of water; $\rho_i$ is density of ice), and $n$ is Glen's n \citep{glen1955} (usually $n=3$).
The first term on the equation's right side represents the opening of the channel through frictional heating, while the following term represents the creep closure of the channel from ice deformation.

@@

The first experiment aims to characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels across a range of channels slopes and  shapes, along with friction factors.
Additionally, we examine the effects of different water discharge variability by smoothing the water discharge over different periods.
The results below present water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and width-integrated shear stress ($w_g\tau_g$, $w_f\tau_f$) from the subglacial and subaerial models.
For brevity, these values together are referred to as ``model outputs'' in the text.
The subglacial model outputs are generally larger than their subaerial equivalents due to their large hydraulic gradients, thus greater sediment transport capacity would be expected in subglacial channels compared to subaerial ones \citep{alley1997}.
To evaluate variability, we subtract the model outputs from their values smoothed using daily averages,  detrending the seasonal signal (Figure~\ref{fig:Qw}).
We then calculate the variability in the model output by taking the standard deviation of this detrended time series.

The effect of channel shape, gradient and roughness is established by running the model with random parameter values of channel slope and geometry factors ($\beta$,$\nabla z_c$ and $\alpha$) and friction factors ($f_r$ and $f_f$; see Table~\ref{table:vpm} for value range).
Runs with parameter combinations are accepted if their mean subglacial water velocity over the season lies between $0.5$\,\unit{m}\,\unit{s}$^{-1}$ and $2$\,\unit{m}\,\unit{s}$^{-1}$ or if subaerial water velocity lies between $0.3$\,\unit{m}\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}\,\unit{s}$^{-1}$, following measurements \citep[e.g.][]{werder2010b,magnusson2012,chandler2013}.
Additionally, to be accepted, subglacial model runs can only experience a flotation fraction ($\frac{\Delta h}{2\,h_o}$) of above $1.2$ for less than $2.5$\,\% of the run.
The initial condition of cross-sectional area $S_g$ is evaluated by applying a model spinup with repeatedly applying the maximum observed water discharge of the first $4$ days of the study until there is no change in the channel area.

@@

Sediment discharge is given by the MPM, EH and Bagnold formulations as
\begin{equation}
  Q_s \propto w\, \tau^{3/2}, \quad Q_s \propto w\, \tau^{5/2}, \quad Q_s \propto w \left(\frac{\Omega}{w}\right)^{3/2} H^{-2/3},
\end{equation}
respectively, for conditions well above the transport onset threshold.

As in Section~\ref{sect:fluv}, the  subaerial channel is assumed to have a width $w$ much greater than its depth $H$, such that $D_h\approx 4h$, and to have a constant head gradient ($\Psi$) given by the topography.
Further, it is  assumed that its width can be approximated by a relation $w \propto Q^\alpha$ (Equation~\ref{eq:wcf}) with $\alpha\in [0,1]$; typically $\alpha \approx 1/2$ (a so-called regime channel).
End members $\alpha=0$ or $\alpha=1$ correspond a  subaerial channels of constant width (a slot canyon) or depth (no natural equivalent), respectively.
% 
For a steady state R-channel, it is assumed that  $\Psi$ is constant (approximated by the gradient of the Shreve \citeyear{shreve1972} potential and that $S$ adjusts such that it is in steady state with $\Psi$ and $Q$.
Note that the R-channel model presented above in Section~\ref{sect:sub_mode} Equation~\ref{eq:dS_dt} calculates $\Psi$ from the time evolving $S_g$ via the Darcy-Weisbach Equation~\ref{eq:dh} and thus no Shreve approximation is then needed.
% Conversely, pipe flow has $S$ fixed and $\Psi$ adjusts to satisfy the given $Q$.
% 
Pipe flow like conditions occur when a R-channel is subjected to rapid discharge variations such that the channel cannot adjust its size significantly on the time scale of the discharge variations.
In this case, it is assumed that the cross-sectional area $S$ is fixed, and $\Psi$ adjusts to the specified $Q$.
For both R-channel and pipe flow, it is assumed that $D_h \propto S^{1/2}$.

With these assumptions the Darcy-Weisbach equation~\eqref{eq:DW} can be solved for the not-fixed quantity: $H$ for a  subaerial channel, $S$ for a R-channel, and $\Psi$ for pipe-flow.
Then, using equations~\eqref{eq:tau-omega}, the shear stress and stream power can be calculated.
These results are summarised in Table~\ref{tab:eqs1}.
Now width integrated shear stress and sediment transport can be calculated for all regimes and for all transport formulas.
The results of the 12 combinations are presented in Table~\ref{tab:Qs} and also in Table~\ref{tab:Qs} where the fractions in the exponents are approximately given by decimal numbers for ease of comparison.

\begin{table}[hbt!]
  \caption{Relations for hydraulic variables for the three drainage regimes:  subaerial channels, R-channel and pipe-flow.  Darcy-Weisbach equation is abbreviated with ``D-W'', and stream power with ``Stream p.''. }
  \small
  \label{tab:eqs1}
  \begin{tabular}{llllll}
    Regime & Fixed & Determined via D-W
    & Additional relations & Shear stress & Stream p.\\
           & & (Equation~\ref{eq:DW})  &  &  \(\tau \propto\) & \(\Omega \propto\)\\
    \hline
    Subaerial & \(\Psi\) & \(H \propto f^{1/3}\, Q^{2/3-2\alpha/3} \, \Psi^{-1/3}\) & \makecell{\(w\,\propto Q^\alpha\) \\ \(S=wH\) \\ \(D_h\propto H\)} & \(f^{1/3} Q^{2/3-2\alpha/3}  \Psi^{2/3}\) & \(Q\, \Psi\)\\
    R-channel & \(\Psi\) & \(S\, \propto f^{2/5}\, Q^{4/5} \, \Psi^{-2/5}\) & \(D_h\propto w \propto H \propto S^{1/2}\) & \(f^{1/5} Q^{2/5} \, \Psi^{4/5}\) & \(Q\, \Psi\)\\
    Pipe & \(S\) & \(\Psi \propto f \, Q^2\, S^{-5/2}\) & \(D_h\propto w \propto H \propto S^{1/2}\) & \(fx Q^2 S^{-2}\) & \(f\, Q^3 S^{-5/2}\)\\
  \end{tabular}
\end{table}

Table~\ref{tab:Qs} shows how the proxy $w\tau$ as well as $Q_s$ of the MPM, EH and Bagnold sediment transport formulas scale with respect to $Q$, $\Psi$ or $S$ for the three different regimes and for different $\alpha$.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow regime.
The total transport formula EH gives a slightly stronger dependence on all variables as should be expected due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$  for the others.
Albeit, the sediment transport response in pipe-flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxies $w_g\,\tau_g$ \& $w_f\,\tau_f$ used below scale only as $Q^2$ for pipe-flow, whereas the transport scales at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{2}$ (a value appropriate for most streams), the exponent on $Q$ for EH is only slightly above what the other transport relations give.

@@

Furthermore, R-channels will rarely operate in steady state as variations in discharge, in particular on the diurnal timescale or the timescale of severe rain or melting events, are too fast for such a channel to reach a steady state.
In such cases they operate more like a pipe of fixed cross-section \citep[e.g.][]{gimbert2016}.
Table~\ref{tab:Qs} shows that in such a situation, the sediment transport scales much more severely with discharge, with the exponent on $Q$ being between $3$ and $5$, compared to the other two regimes when that exponent is at most $1.7$ (Table~\ref{tab:Qs}).
Thus fluctuations of discharge on short timescales (on the order of a day) have the potential to cause conditions with very high sediment transport capacities, that are of far high magnitude than the subaerial channels.

@@

\FloatBarrier
\subsection{Increased variability in evolving subglacial channels occurs across a range of channel shapes, slopes  and friction values}
\label{sect:ensemble}

The first numerical experiment aims to compare the variability between the subglacial and subaerial  model outputs to a range of channel shapes, friction factors, and water discharge variability, as the subglacial channel's cross-sectional area is allowed to evolve.
Following the findings in Section~\ref{sect:scaling}, variability in subglacial sediment discharge capacity could approach that of subaerial systems, should water discharge variations be sufficienlty small or subglacial channels respond in size quickly.
For simplicity, we only examine $w\tau$, not the sediment transport formulas mentioned above.
%We have chosen to exclude the pipe flow case (Section~\ref{sect:scaling}) from analysis as it depends greatly on the choice of the subglacial cross sectional area $S_g$.

Across the range of parameters  examined (Section\,\ref{sect:imp}), variability in all model outputs remains higher in the subglacial system compared to the subaerial one for both \alpine{} and \icesheet{} (Figure~\ref{fig:multi_run}).
In some cases, the variability in subglacial model outputs is a  magnitude larger than their subaerial counterparts.
Results from  both \alpine{} and \icesheet{} scenarios show that variability in both subglacial and subaerial outputs decreases with smoothing time longer than approximately 1--5 days (Figure~\ref{fig:multi_run}), when the diurnal variations in water discharge are removed from smoothing.
In the \icesheet case, the smaller variability in model outputs compared to \alpine may in fact mean that the R-channel is in closer equilibrium between channel size and water discharge (Section~\ref{sect:scaling}).
In this case, water discharge variations would create smaller disparities in sediment transport capacity than even in subaerial cases as the subglacial channel size evolves to accommodate them (Table~\ref{tab:Qs}).

@@

With respect to the  manuscript's first objective,
results demonstrate that sediment transport capacity variability is higher in subglacial channels with evolving size across a range of channel shapes and friction factors.
Increased variability even occurs in environments where the water discharge has relatively small diurnal variations, as in the \icesheet case.
The role of this increased noise in sediment transport capacity in evaluating hydro-climatic signals from sediment records in glacierized catchments needs further evaluation.
Furthermore, this processes may lead to sediment exhaustion by subglacial water repeatedly crossing the mobilization threshold across a range of water discharges.

>>>>

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom, which varies with water velocity and channel width.
  In subaerial channels, water discharge variations are accommodated by flow depth and width changes, along with water velocity.
  However, in subglacial channels,  water discharge changes mainly result in velocity variations, as the channel geometry evolves slowly (over days).
  Here, we present relationships between water discharge and sediment transport under several channel types.
  They demonstrate that sediment transport is highly non-linear with respect to water discharge with fixed subglacial channel size, and in fact, can very less than subaerial flow if water discharge changes slowly enough to be in equilibrium with subglacial channel size.
  We then apply hydrographs from an Alpine glacier and the Greenland Ice Sheet to subglacial and subaerial hydraulics models, revealing greater variability in the inferred sediment transport capacity in subglacial than in subaerial channels, as subglacial channels evolve in size.
  This variability stems from the slowly evolving channel size that forces water velocity to accommodate variations in water discharge.
  Furthermore, the changing channel size means that sediment transport capacity peaks before the maximum water discharge.
  Results also show that high subglacial sediment transport rates and water velocities can occur across a wide range of water discharges, meaning that in a transport-limited system, there could be a variable relationship between sediment and water discharge.
  These findings are discussed in the context of understanding sediment discharge from glaciers with different hydro-climatic forcings and assumptions of channel shape.
  We also examine the implications for interpreting records of sediment discharge in glacierized catchments
\end{abstract}

@@

Periods of millennia or longer, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \citep[c.f.][]{hallet1979,iverson2012,ugelvig2018}.
Pressurized subglacial water can transport this sediment from underneath glaciers \citep{walder1994,creyts2013,beaud2018,delaney2019} if it is reachable by the water.

@@

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.6\linewidth]{Fig1.pdf}
  \caption{Sketch for the different responses of subglacial and subaerial channels to increased water discharge over short time scales.
    Arrow length denotes water velocity magnitudes in the subglacial (subaerial) $v_g$ ($v_f$) channels.
    $S_g$ ($S_f$) represents the wetted area in subglacial (subaerial) channels.
    The subglacial channel width $w_g$ remains unchanged, while the subaerial channel width $w_f$ evolves with water discharge.
    Subglacial (subaerial) shear stress $\tau_g$ ($\tau_f$) is responsible for the mobilization of sediment.
  }
  \label{fig:cartoon}
\end{figure}

Sediment mobilization in subaerial and subglacial channels, therefore, responds differently to changing water discharge.
These differences are implicitly included in a range of models quantifying sediment transport in both subglacial and subaerial channels  \citep[e.g.][]{walder1994,alley1997,tucker1997,creyts2013,beaud2018,delaney2019,hewitt2019,wickert2019}.
To date, explicit parameterizations of subglacial sediment transport capacity with respect to water discharge have assumed fixed channel size and demonstrate a very non-linear response in sediment transport capacity to water discharge \citep{alley1997}.
The modeling frameworks examining subglacial sediment transport with evolving channel size minimally discuss the impact of different hydrological forcing on sediment transport capacity to understand the result of hydro-climatic conditions on subglacial sediment dynamics \citep{creyts2013,beaud2018,delaney2019,hewitt2019}.
Yet, sediment discharge capacity controls the mobilization and deposition of sediment, making it one of the most fundamental aspects of subglacial sediment evacuation, especially over short timescales.

Despite the physical differences between the subaerial and subglacial systems, water discharge and sediment export are often compared in contemporary glacial environments \citep[e.g.][]{willis1996,pearce2003,richards2003,swift2005,chu2009,tedstone2012,chu2012,overeem2017,delaney2018,swift2021,lu2022,andresen2024}.
These comparisons are natural given that the majority of sediment evacuated from glaciers over contemporary seasonal timescales is done so fluvially \citep[e.g.][]{collins1989, riihimaki2005,delaney2024}.
In subaerial systems, water discharge and sediment transport capacity for a grain size are closely linked \citep[e.g. ][]{schmidt2006,pitlick2021} \ian{check citiation}.
However, the potential for a highly variable response in sediment transport capacity in subglacial sediment transport may affect the interpretation of sediment transport records in glacial systems \citep[e.g.][]{ganti2016,mancini2023}, if properties associated with subaerial systems are assumed.
This relationship should be clarified as sediment dynamics change in glacierized regions along with changing hydrology, increased variability of water discharge \citep{lane2019} and extreme glacier melt  \citep[e.g.][]{cremona2023,overeem2015} that may drive high magnitude sediment transport events.

This manuscript has two objectives: 1) evaluate the variability of subglacial sediment transport capacity with evolving subglacial channel size and different hydrographs and 2) establish if sub-seasonal variations in water discharge can co-vary with sediment transport capacity in subglacial systems.
First, to reinforce the highly variable behavior of subglacial sediment transport capacity,  we present algebraic formulations of the sediment transport capacity response to water discharge in subaerial, pipe-flow, and equilibrium R-channel conditions, extending the relationships presented in \citet{alley1997}.
We then use numerical models to examine sediment transport capacity in subglacial channels that evolve in size and subaerial channels.
They are then applied to proglacial hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and a land-terminating glacier in Greenland (Leverett Glacier).
We run models with an ensemble of parameters to evaluate the sediment transport capacity variability in subglacial channels compared to subaerial ones.
Lastly, model outputs demonstrate the particular processes that can drive variability in sediment discharge capacity in subglacial systems.
Together, these findings indicate differences in the relationships between water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
The manuscript then uses the identified differences between the subaerial and subglacial channel types to discuss the implications for interpreting the processes driving variability in sediment transport records and the impact of these differences on sediment export from glaciers.

@@

\section{Methods}
\label{sect:meth}
The two models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table~\ref{table:vpm}).
Both models use the measured discharge to calculate water velocity, shear stress, and width-integrated shear stress, upon which both suspended sediment and bedload transport depend \citep[Figure \ref{fig:cartoon}; ][]{shields1936}.
Our evaluation in terms of shear stress omits the selection of a sediment transport relationship and a grain-size parameter \citep[e.g.][]{shields1936,meyer1948}.

@@

Here, it is assumed that the water is transported through a subglacial channel \citep[Figure~\ref{fig:cartoon}; ][]{rothlisberger1972} beneath a glacier with channel length $l$, with a flatbed and a mean thickness of $h_{ice}$.
The channel size grows from melt due to frictional heating from water flow and closes due to the creep of the ice.
The formulation here does not consider the englacial storage of water.
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_1 \frac{Q \Delta h}{l} - C_2 \left(h_{o}-\frac{\Delta h}{2}\right)^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent where $t$ is time, $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (complete values given in Table~\ref{table:vpm}), $g$ is the acceleration due to gravity, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change over $l$, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the mean ice overburden pressure expressed in meter water equivalent ($\rho_w$ is density of water; $\rho_i$ is density of ice), and $n$ is Glen's n \citep{glen1955} (usually $n=3$).
The first term on the equation's right side represents the opening of the channel through frictional heating, while the following term represents the creep closure of the channel from ice deformation.

@@

The first experiment aims to characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels across a range of channel slopes and shapes, along with friction factors.
Additionally, we examine the effects of different water discharge variability by smoothing the water discharge over different periods.
The results below present water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and width-integrated shear stress ($w_g\tau_g$, $w_f\tau_f$) from the subglacial and subaerial models.
For brevity, these values together are referred to as ``model outputs'' in the text.
The subglacial model outputs are generally larger than their subaerial equivalents due to their large hydraulic gradients, thus greater sediment transport capacity would be expected in subglacial channels compared to subaerial ones \citep{alley1997}.
To evaluate variability, we subtract the model outputs from their values smoothed using daily averages,  detrending the seasonal signal (Figure~\ref{fig:Qw}).
We then calculate the variability in the model output by taking the standard deviation of this detrended time series.

The effect of channel shape, gradient, and roughness is established by running the model with random parameter values of channel slope and geometry factors ($\beta$,$\nabla z_c$, and $\alpha$) and friction factors ($f_r$ and $f_f$; see Table~\ref{table:vpm} for value range).
Runs with parameter combinations are accepted if their mean subglacial water velocity over the season lies between $0.5$\,\unit{m}\,\unit{s}$^{-1}$ and $2$\,\unit{m}\,\unit{s}$^{-1}$ or if subaerial water velocity lies between $0.3$\,\unit{m}\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}\,\unit{s}$^{-1}$, following measurements \citep[e.g.][]{werder2010b,magnusson2012,chandler2013}.
Additionally, to be accepted, subglacial model runs can only experience a flotation fraction ($\frac{\Delta h}{2\,h_o}$) of above $1.2$ for less than $2.5$\,\% of the run.
The initial condition of cross-sectional area $S_g$ is evaluated by applying a model spinup with repeatedly applying the maximum observed water discharge of the first $4$ days of the study until there is no change in the channel area.

@@

Sediment discharge is given by the MPM, EH, and Bagnold formulations as
\begin{equation}
  Q_s \propto w\, \tau^{3/2}, \quad Q_s \propto w\, \tau^{5/2}, \quad Q_s \propto w \left(\frac{\Omega}{w}\right)^{3/2} H^{-2/3},
\end{equation}
respectively, for conditions well above the transport onset threshold.

As in Section~\ref{sect:fluv}, the subaerial channel is assumed to have a width $w$ much greater than its depth $H$, such that $D_h\approx 4h$, and to have a constant head gradient ($\Psi$) given by the topography.
Further, it is assumed that its width can be approximated by a relation $w \propto Q^\alpha$ (Equation~\ref{eq:wcf}) with $\alpha\in [0,1]$; typically $\alpha \approx 1/2$ (a so-called regime channel).
End members $\alpha=0$ or $\alpha=1$ correspond a  subaerial channels of constant width (a slot canyon) or depth (no natural equivalent), respectively.
% 
For a steady state R-channel, it is assumed that  $\Psi$ is constant (approximated by the gradient of the Shreve \citeyear{shreve1972} potential and that $S$ adjusts such that it is in steady state with $\Psi$ and $Q$.
Note that the R-channel model presented above in Section~\ref{sect:sub_mode} Equation~\ref{eq:dS_dt} calculates $\Psi$ from the time-evolving $S_g$ via the Darcy-Weisbach Equation~\ref{eq:dh} and thus no Shreve approximation is then needed.
% Conversely, pipe flow has $S$ fixed and $\Psi$ adjusts to satisfy the given $Q$.
% 
Pipe flow-like conditions occur when an R-channel is subjected to rapid discharge variations such that the channel cannot adjust its size significantly on the time scale of the discharge variations.
In this case, it is assumed that the cross-sectional area $S$ is fixed, and $\Psi$ adjusts to the specified $Q$.
For both R-channel and pipe flow, it is assumed that $D_h \propto S^{1/2}$.

With these assumptions, the Darcy-Weisbach equation~\eqref{eq:DW} can be solved for the not-fixed quantity: $H$ for a subaerial channel, $S$ for a R-channel, and $\Psi$ for pipe-flow.
Then, using equations~\eqref{eq:tau-omega}, the shear stress and stream power can be calculated.
These results are summarised in Table~\ref{tab:eqs1}.
Now width integrated shear stress and sediment transport can be calculated for all regimes and all transport formulas.
The results of the 12 combinations are presented in Table~\ref{tab:Qs} and also in Table~\ref{tab:Qs} where the fractions in the exponents are approximately given by decimal numbers for ease of comparison.

\begin{table}[hbt!]
  \caption{Relations for hydraulic variables for the three drainage regimes:  subaerial channels, R-channel, and pipe-flow.  Darcy-Weisbach equation is abbreviated with ``D-W'', and stream power with ``Stream p.''. }
  \small
  \label{tab:eqs1}
  \begin{tabular}{llllll}
    Regime & Fixed & Determined via D-W
    & Additional relations & Shear stress & Stream p.\\
           & & (Equation~\ref{eq:DW})  &  &  \(\tau \propto\) & \(\Omega \propto\)\\
    \hline
    Subaerial & \(\Psi\) & \(H \propto f^{1/3}\, Q^{2/3-2\alpha/3} \, \Psi^{-1/3}\) & \makecell{\(w\,\propto Q^\alpha\) \\ \(S=wH\) \\ \(D_h\propto H\)} & \(f^{1/3} Q^{2/3-2\alpha/3}  \Psi^{2/3}\) & \(Q\, \Psi\)\\
    R-channel & \(\Psi\) & \(S\, \propto f^{2/5}\, Q^{4/5} \, \Psi^{-2/5}\) & \(D_h\propto w \propto H \propto S^{1/2}\) & \(f^{1/5} Q^{2/5} \, \Psi^{4/5}\) & \(Q\, \Psi\)\\
    Pipe & \(S\) & \(\Psi \propto f \, Q^2\, S^{-5/2}\) & \(D_h\propto w \propto H \propto S^{1/2}\) & \(fx Q^2 S^{-2}\) & \(f\, Q^3 S^{-5/2}\)\\
  \end{tabular}
\end{table}

Table~\ref{tab:Qs} shows how the proxy $w\tau$ as well as $Q_s$ of the MPM, EH, and Bagnold sediment transport formulas scale with respect to $Q$, $\Psi$ or $S$ for the three different regimes and for different $\alpha$.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow regime.
The total transport formula EH gives a slightly stronger dependence on all variables as should be expected due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$  for the others.
Albeit, the sediment transport response in pipe flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxies $w_g\,\tau_g$ \& $w_f\,\tau_f$ used below scale only as $Q^2$ for pipe-flow, whereas the transport scales at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts the relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{2}$ (a value appropriate for most streams), the exponent on $Q$ for EH is only slightly above what the other transport relations give.

@@

Furthermore, R-channels will rarely operate in a steady state as variations in discharge, in particular on the diurnal timescale or the timescale of severe rain or melting events, are too fast for such a channel to reach a steady state.
In such cases they operate more like a pipe of fixed cross-section \citep[e.g.][]{gimbert2016}.
Table~\ref{tab:Qs} shows that in such a situation, the sediment transport scales much more severely with discharge, with the exponent on $Q$ being between $3$ and $5$, compared to the other two regimes when that exponent is at most $1.7$ (Table~\ref{tab:Qs}).
Thus fluctuations of discharge on short timescales (on the order of a day) have the potential to cause conditions with very high sediment transport capacities, that are of far higher magnitude than the subaerial channels.

@@

\FloatBarrier
\subsection{Increased variability in evolving subglacial channels occurs across a range of channel shapes, slopes,  and friction values}
\label{sect:ensemble}

The first numerical experiment aims to compare the variability between the subglacial and subaerial model outputs to a range of channel shapes, friction factors, and water discharge variability, as the subglacial channel's cross-sectional area is allowed to evolve.
Following the findings in Section~\ref{sect:scaling}, variability in subglacial sediment discharge capacity could approach that of subaerial systems, should water discharge variations be sufficiently small or subglacial channels respond in size quickly.
For simplicity, we only examine $w\tau$, not the sediment transport formulas mentioned above.
%We have chosen to exclude the pipe flow case (Section~\ref{sect:scaling}) from analysis as it depends greatly on the choice of the subglacial cross sectional area $S_g$.

Across the range of parameters examined (Section\,\ref{sect:imp}), variability in all model outputs remains higher in the subglacial system compared to the subaerial one for both \alpine{} and \icesheet{} (Figure~\ref{fig:multi_run}).
In some cases, the variability in subglacial model outputs is a magnitude larger than their subaerial counterparts.
Results from both \alpine{} and \icesheet{} scenarios show that variability in both subglacial and subaerial outputs decreases with smoothing time longer than approximately 1--5 days (Figure~\ref{fig:multi_run}), when the diurnal variations in water discharge are removed from smoothing.
In the \icesheet case, the smaller variability in model outputs compared to \alpine may mean that the R-channel is in closer equilibrium between channel size and water discharge (Section~\ref{sect:scaling}).
In this case, water discharge variations would create smaller disparities in sediment transport capacity than even in subaerial cases as the subglacial channel size evolves to accommodate them (Table~\ref{tab:Qs}).

@@

Concerning the manuscript's first objective,
results demonstrate that sediment transport capacity variability is higher in subglacial channels with evolving size across a range of channel shapes and friction factors.
Increased variability occurs in environments where the water discharge has relatively small diurnal variations, as in the \icesheet case.
The role of this increased noise in sediment transport capacity in evaluating hydro-climatic signals from sediment records in glacierized catchments needs further evaluation.
Furthermore, this process may lead to sediment exhaustion by subglacial water repeatedly crossing the mobilization threshold across a range of water discharges.
